Relationships in the extremes and their influence on competition and coexistence

### 2021-09-29 paper/Paper.Rnw
finished editing for now, through the Results, but no Discussion yet

\affil[a]{Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas}
\affil[b]{Laboratory of Populations, Rockefeller University}

@@

%***Add to these
\keywords{competition, coexistence, biodiversity, modern theory of coexistence, storage effects, asymmetric tail association}

%Here are the paragraph-level "rules" of writing, for reference: 
%Every paragraph needs to have a clear topic sentence.
%It needs to be one of the first three sentences.
%If it is not the first sentence, then the sentences preceding it are for background, in order to make it possible to state the topic sentence.
%The topic sentence is the main *result* or *claim* of the paragraph, not just a statement of what the topic is.
%All sentences in the paragraph support and validate the claim made in the topic sentence.

@@

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as `relative nonlinearity
in competition' and `storage effects' \citep{Chesson_1994} which will be familiar to many readers. The MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although the MTC is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in figure \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also used, in addition to correlations
(e.g., various common forms of regression), but correlations are extremely common, and other common
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in figure \ref{fig:noise}A,C 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (figure \ref{fig:noise}A, termed ``left-tail association'') or right (figure \ref{fig:noise}C,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

%Intro to storage effects and why they might be influenced by ATAs
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. 
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
%This is reasonable assumption for non-rare species, since 
%a better environment leads to greater numbers which leads to more intraspecific competition, but a rare species may be essentially unaffected
%by intraspecific competition. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor; dormant stages such as are in seed banks are an
example buffering mechanisms \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional reason to quantify the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can or may alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive, potentially 
influencing spatial coexistence mechanisms.

%make the above abstract ideas more concrete
To make the above ideas more concrete, we briefly mention examples from prior work to 
which the ideas may apply. 
\cite{Chesson_2012} provides two examples of competitive plant communities in which different species
are likely influenced differently by phenologically distict environmental variables: an Australian 
grass-forb community, and an Arizona community of desert annuals. 
The communities follow the criteria required for temporal storage effects to operate \citep{Chesson_2012}. In both 
examples, distinct species have different periods of germination, competition is influenced by resource 
consumption, and population growth is buffered by overlapping generations or via seed banks.
Temporal differences between species in germination rates suggest the possibility that 
population growth of distinct species may be driven by rainfall signals in distinct parts
of the growing season. Thus all the ingredients seem to be present by which climatic changes may
alter ATAs between environmental variables which are relevant for the coexistence (or not)
of species, motivating our efforts, here, to perform a detailed study of how ATAs can influence competition.

%***JAS: I am trying to relay the message that climate change and specifically extreme events have consequences for coexistence and competition
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??

%***DAN: Do it like you have it but shorter.
%the RAMPAGE of extreme events such as heat-waves, floods, hurricanes, fires, etc. implications for coexistence and competition. 
%***DAN: Possible topic sentence like thing after talking to Jasmin on 9/17: 
Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied, the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, have been studied in several systems.  
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}. While both changes can be attributed to extreme events, shifts in mean are more-so a result of gradual trend effects, and increased variance is a result of extreme event effects \citep{Jentsch_2007}.
However, impacts on competition and coexistence of climate change have been more commonly researched via studying changes in the mean of environmental variables. 
%Jasmin tried to split this para up into effects on coexistence of changes in mean, starting here...
Decreased mean growth season precipitation, for instance, can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}, and coexistence is predicted to collapse between Mediterranean tree species in accordance to forecasted mean rainfall level decrease \cite{diPaola_2012}. 
Alternatively, dominant species exclusion may be beneficial as rare species can increase in diversity \citep{Mariotte_2013}, while an increase in dominant species fitness can reduce community diversity, typical of invasive species \citep{Vitousek_1997}. 
As a product of changing means, range shifts can favor plant coexistence by introducing new herbivores \cite{Descombes_2020} or disfavor coexistence through increased competition \cite{Usinowicz_2021}.

%...and variance...
While the body of literature on the impact of increased environmental variability has expanded, event-based studies accounted for only one-fifth of experimental climate change research as of 2006 \citep{Jentsch_2007}, the most commonly manipulated extremes being temperature and rainfall.
Despite research on climate variability in relation to competition and coexistence being only a portion of that, we review some studies here.
Extreme drought and heat waves have shown to impact competitive output, rearranging species composition and dominance of grass-forb communities \citep{White_2000, White_2001, Hoover_2014}; although complete exclusion of dominant species is uncommon \citep{Lloret_2012}. \cite{Fay_2003} experimentally manipulated rainfall for more frequent and intense precipitation while keeping constant the total volume of rain over Kansas tallgrass prairie and found that while some species had reduced fitness, overall community structure remained the same. For invasive species, increased climate and resource variability allowed for greater success of invasibility in terms of density and competitive equilibrium \citep{Parepa_2013}. 
%...and extreme events (though this blends together with changes in variance)
Although increased variability of environmental fluctuations is a newer consideration for studying environmental changes, environmental variability has been analytically shown to be important for classical coexistence, i.e. MCT \citep{Chesson_1981, Warner_1985}. Moreover, by using MCT, increased variability has been shown to promote coexistence in plants communities, via the aforementioned storage effect \citep{Adler_2006,  Angert_2009}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence. 
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general, they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).
%DAN: So what jasmin was trying to do was summarize a lot of lit on effects of climate change on coexistence, in these three parts
%(means, varinace, and extreme events).

%Move upl the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, 
hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. 
We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year.
For each $i$, we assume the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
However, we do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. We will consider various distributions of $\left(b_1, b_2\right)$ with the above 
properties, corresponding to various tail association cases. Note that $N$ is constant through time.
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
We simulated two different communities up to \Sexpr{M} time steps to hopefully make it clear to the that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog} for more details). The only distinction between the two simulations is that one depends on asymmetrically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverability when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

%Goal para
We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. Need better ending sentence.

@@

We here pursue goal G1 from the Introduction, to augment the MTC to show how ATA
plays a role. For pedgogical clarity, we develop each step of the theory in general and for 
a simple example model, the lottery model (see \ref{section:introduction}).

%***DAN: Note the inclusion of the model here means we will have to edit the "Model" section of sup mat, since we don't need to intro the model twice. Do later.

Modern coexistence theory \citep{Chesson_2000} and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
different mechanisms to species coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
We now define storage effects for a general two species model, and specifically 
for the lottery model, using an approach and notation
similar to those of \cite{Ellner_2016,Ellner_2019}; see those papers for additional details.

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. As for the lottery model example, we here assume, for simpicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

@@

For the lottery model, recall that $E_i=B_i$ and $C_{i \bs i} = B_j/\delta$. So 
$\varphi^{-1} \circ F_{E_i}(E_i) = \frac{b_i-\mu_i}{\sigma}$ and 
$\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})=\frac{b_j - \mu_j}{\sigma}$.  
Also, $\rho_i = \cov(\frac{b_i-\mu_i}{\sigma},\frac{b_j-\mu_j}{\sigma})$, which equals the 
Pearson correlation of $b_i$ and $b_j$. For the lottery model, it seems natural to make the choice
$\ddot{\rho}_i=\rho_i$, which straightforwardly implies that $(E_i^\nat,C_{i \bs i}^\nat)$
can be written as $(\exp(\ddot{b}_i),\exp(\ddot{b}_j))$, where $(\ddot{b}_i,\ddot{b}_j)$
is bivariate-normally distributed with mean $(\mu_i,\mu_j)$, $\var(\ddot{b}_i)=\var(\ddot{b}_j)=\sigma$,
and $\cov(\ddot{b}_i,\ddot{b}_j)=\cov(b_i,b_j)$. So $(\ddot{b}_i,\ddot{b}_j)$ has the same
marginal distributions and covariance matrix as $(b_i,b_j)$, but it is bivariate normal, whereas
$(b_i,b_j)$ itself need not be. Given these definitions, it is straightforward to compute the 
component $\olrn_{i \bs i}$ of (\ref{eq:storage_effects_further1}) for the lottery model using the simulation 
approach of \cite{Ellner_2016,Ellner_2019}. Note that $E_j$ and $C_{j \bs i}$ are linearly related for the lottery model,
so $E_j^\nat=E_j$, $C_{j \bs i}^\nat = C_{j \bs i}$, and $\olrn_{j \bs i}=\olr_{j \bs i}$.

>>>>

\affil[a]{Department of Ecology and Evolutionary Biology and Center for Ecological Research and Kansas Biological Survey, University of Kansas}
\affil[b]{Laboratory of Populations, Rockefeller University}

@@

%***Add to these
\keywords{asymmetric tail association, biodiversity, coexistence, competition, modern coexistence theory, storage effects}

@@

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other common
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}A,C 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}A, termed ``left-tail association'') or right (Fig. \ref{fig:noise}C,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

%Intro to storage effects and why they might be influenced by ATAs
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. 
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simulations using a simple two-species model, the lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is then the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N$ is constant through time. 
For each $i$, we assume for simplicity that the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
We do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. 
For proof of concept, we simulated two communities for \Sexpr{M} time 
steps each, using symmetric and asymmetric tail associations (Fig. \ref{fig:noise}) between $b_1$ and $b_2$, respectively, in the
two simulations. Fig. \ref{fig:pedagog} gives details of the 
simulations and shows that ATAs substantially influenced aspects of species coexistence for this model setup.

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can or may alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive.

%make the above abstract ideas more concrete
To make the above ideas more concrete, we briefly mention examples from prior work to 
which the ideas may apply. 
\cite{Chesson_2012} provides two examples of competitive plant communities in which different species
are likely influenced differently by phenologically distict environmental variables: an Australian 
grass-forb community, and an Arizona community of desert annuals. 
The communities follow the criteria required for temporal storage effects to operate \citep{Chesson_2012}. In both 
examples, distinct species have different periods of germination, competition is influenced by resource 
consumption, and population growth is buffered by overlapping generations or via seed banks.
Temporal differences between species in germination rates suggest the possibility that 
population growth of distinct species may be driven by rainfall signals in distinct parts
of the growing season. Thus all the ingredients seem to be present by which climatic changes may
alter ATAs between environmental variables which are relevant for the coexistence
of the example species, motivating our efforts, here, to perform a detailed study of how ATAs can influence competition.

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence
Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
not been studied, the importance for coexistence of other aspects of climate change, including changes 
in means and measures of variability of environmental variables, has been studied in many systems, so we briefly 
review some of that large, related literature.
For instance, \cite{diPaola_2012} concluded that coexistence of drought sensitive,
flood resistant species with flood sensitive, drought resistant species in 
Mediteranean coastal swamp forests of central Italy would be jeopardized
by projected changes in mean rainfall for the region. In an interesting example
involving climate-change effects on herbivore-mediated competition in plants,
\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
higher altitudes of herbivore populations than of plant populations, newly arrived
herbivores at high altitudes can inflate species richness by reducing dominant plant species 
in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
all species they considered, but that changes in competition would more than offset these direct benefits,
and, as a result, some species would be extirpated. Numerous studies also take an ``event-focussed'' 
perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
form of extreme weather events, can alter competitive interactions. For instance, 
\cite{White_2001} showed that single extreme heating events can alter competition 
between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
nutrient availability for plants, which is mediated by soil moisture content, can alter 
the potential for invasive species to enter an ecosystem and establish, a process that involves
the invader overcoming competition from resident species. 
Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
communities, via storage effects \citep{Adler_2006,  Angert_2009}.
Changes in variability and in extreme events are distinct from the changes in ATAs we 
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

%***DAN: Jasmin, I commented the below sentence because that body of work is famously 
%complex and sophisticated, and I did not feel we could do it justice in short space.
%decreased growth season precipitation can decrease the competitive ability of dominant 
%forest tree species \citep{Clark_2011}, 
%***DAN: Jasmin, I commented the below sentence because I'm not sure the two studies 
%cited there are actually directly about coexistence, in the MCT sense. One of them is
%a general review on invasive species, and the other has to do with competition,
%but coexistence is not mentioned. I searched both papers for "coexist" and found
%no hits. OK, both are relevant, but maybe not so directly, to what we are talking about,
%and since this paragraph could anyway stand to be shortened a bit, I took these out.
%Alternatively, dominant species exclusion may be beneficial as rare species can increase 
%in diversity \citep{Mariotte_2013}, while an increase in dominant species fitness can 
%reduce community diversity, typical of invasive species \citep{Vitousek_1997}.

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs exist and are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATA plays a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, known mechanisms of coexistence. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical and importance and its importance for real ecosystems.

@@

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. For pedgogical clarity, we develop each step of our new theory for a general 
two-species model, and for the lottery model (Introduction).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
So we begin by defining storage effects for a general two species model and 
for the lottery model. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}.

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. As for the lottery model example (Introduction), we here assume, for simpicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

@@

For the lottery model, recall that $E_i=B_i$ and $C_{i \bs i} = B_j/\delta$. So 
$\varphi^{-1} \circ F_{E_i}(E_i) = \frac{b_i-\mu_i}{\sigma}$ and 
$\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})=\frac{b_j - \mu_j}{\sigma}$.  
Also, $\rho_i = \cov(\frac{b_i-\mu_i}{\sigma},\frac{b_j-\mu_j}{\sigma})$, which equals the 
Pearson correlation of $b_i$ and $b_j$. For the lottery model, it seems natural to make the choice
$\ddot{\rho}_i=\rho_i$, which straightforwardly implies that $(E_i^\nat,C_{i \bs i}^\nat)$
can be written as $(\exp(\ddot{b}_i),\exp(\ddot{b}_j))$, where $(\ddot{b}_i,\ddot{b}_j)$
is bivariate-normally distributed with mean $(\mu_i,\mu_j)$, $\var(\ddot{b}_i)=\var(\ddot{b}_j)=\sigma$,
and $\cov(\ddot{b}_i,\ddot{b}_j)=\cov(b_i,b_j)$. So $(\ddot{b}_i,\ddot{b}_j)$ has the same
marginal distributions and covariance matrix as $(b_i,b_j)$, but it is bivariate normal, whereas
$(b_i,b_j)$ itself was not required to be. Given these definitions, it is straightforward to compute the 
component $\olrn_{i \bs i}$ of (\ref{eq:storage_effects_further1}) for the lottery model using the simulation 
approach of \cite{Ellner_2016,Ellner_2019}. Note that $E_j$ and $C_{j \bs i}$ are linearly related for the lottery model,
so $E_j^\nat=E_j$, $C_{j \bs i}^\nat = C_{j \bs i}$, and $\olrn_{j \bs i}=\olr_{j \bs i}$.
